---
title: "CD8 TIL integration with STACAS"
author: "Paul Gueguen, Massimo Andreatta, and Santiago  Carmona"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

```{r, include=FALSE, fig.width=16, fig.height=12}
library(renv)
renv::restore()
library(Seurat)
library(ggplot2)
library(scGate)
library(STACAS)
library(SignatuR)
library(dplyr)
```

We start from integrated collection of large and high-quality samples, generated using 2-Integration_STACAS.Rmd. Here we integrate the remaining datasets on top of the integrated 'reference' set.

# Set up parameters

```{r}
path_output <- "./out/"
npcs <- 50
nfeatures <- 800
seed <- 123
set.seed(seed)
umap.neighbors <- 30
semisup=TRUE

# Setup color palette
mycols <- c('NaiveLike' = '#b7d2e0', 'CM' = '#da6f6f','EM'= '#72b28a','TEMRA' = '#e5bfaf', 'TPEX' = '#aca6e0' , 'TEX' ='#f5d39f', "MAIT" = '#fdbfd4')

mycols_scGate <- c('CD8_N' = '#b7d2e0', 'CD8_EM'= '#72b28a','CD8_TEMRA' = '#e5bfaf', 'CD8_TPEX' = '#aca6e0' , 'CD8_TEX' ='#f5d39f', "CD8_MAIT" = '#fdbfd4')
```

# Load data

```{r}
seurat.all <- readRDS('cache/Seurat.list.Utility.v0.4_step6.rds')
length(seurat.all)

preintegrated <- readRDS('out/Utility.v0.4.CD8.integrated.stacas.semisup.rds')
```

Parameters
```{r}
npcs <- ncol(preintegrated@reductions$pca)
hvg <- VariableFeatures(preintegrated)
```

Prepare data
```{r}
inref <- unique(preintegrated$SampleLabel)
notinref <- setdiff(names(seurat.all), inref)
seurat.all <- seurat.all[notinref]

#small-scale test
seurat.all <- seurat.all[100:105]
seurat.all <- lapply(seurat.all, NormalizeData)

#may want to exclude small samples
n <- unlist(lapply(seurat.all, ncol))
head(sort(n))
```

Prepare reference sets
```{r}
#copy assay
preintegrated <- RenameAssays(preintegrated, integrated = "preintegrated")
DefaultAssay(preintegrated) <- "preintegrated"
```

# Unsupervised STACAS integration

We will use scGate labels to guide STACAS integration.

```{r}
ref <- list('reference'=preintegrated)
seurat.list <- append(seurat.all, values=ref, after=0)

sapply(seurat.list, DefaultAssay)

stacas_anchors <- FindAnchors.STACAS(seurat.list,
                                     reference = 1,
                                     cell.labels = NULL,
                                     anchor.features = hvg,
                                     dims = 1:npcs)

tree <- SampleTree.STACAS(stacas_anchors, obj.names = names(seurat.list))

integrated <- IntegrateData.STACAS(anchorset=stacas_anchors, dims=npcs,
                                   sample.tree=tree)

```


Reduced dimensionalities
```{r}
integrated <- ScaleData(integrated) |> RunPCA(dims = 1:npcs) |>
      RunUMAP(reduction = "pca", dims = 1:npcs, seed.use=seed)
```

See results
```{r}
a <- DimPlot(integrated, reduction = "umap",
        cols = mycols, group.by = "functional.cluster", raster = T) +
  ggtitle('Annotation') + theme(aspect.ratio = 1)

b <- DimPlot(integrated, reduction = "umap",
        group.by = "SampleLabel", raster = T) +
  theme(aspect.ratio = 1, legend.text = element_text(size=7),
        legend.key.size = unit(0.1, 'cm'))

b | a
```

#Semi-supervised approach

Predict CD8T subtypes using scGate
```{r}
# Setup function to run scGate + seurat pipeline
run_seurat <- function(x, bl=NULL, nfeat=1000, npca=30) {
  
  x <- x %>% NormalizeData(verbose=F) %>%  
      FindVariableFeatures.STACAS(nfeat = nfeat, genesBlockList = bl) %>%
      ScaleData(verbose=F) %>%
      RunPCA(npcs=npca, verbose=F) %>% RunUMAP(dims=1:npca, verbose=F)

  return(x)
}

# define a function using variable object_in
run_scGate_CD8T <- function(x, models) {

  x <- scGate(x, model=models, multi.asNA=T, ncores=8,
           reduction="pca", assay = "RNA")

  return(x)
}
```

```{r}
scGate_models <- scGate::get_scGateDB(branch = "dev")

cd8.til.models <- scGate_models$human$CD8_TIL
cd8.til.models <- cd8.til.models[names(cd8.til.models) %in% c("CD8_Tinn","CD8_TRM") == FALSE]

my.genes.blocklist <- SignatuR::GetSignature(SignatuR$Hs)

names <- names(seurat.all)
seurat.all <- lapply(
  seq_along(seurat.all), 
  function(i) {
    print(sprintf("Dataset %i of %i", i, length(seurat.all)))
    x <- seurat.all[[i]]
    x <- run_seurat(x, bl=my.genes.blocklist, nfeat=1000, npca=30)
    run_scGate_CD8T(x, models=cd8.til.models)
  })
names(seurat.all) <- names

```

```{r}
ref <- list('reference'=preintegrated)
seurat.list <- append(seurat.all, values=ref, after=0)

sapply(seurat.list, DefaultAssay)

stacas_anchors <- FindAnchors.STACAS(seurat.list,
                                     reference = 1,
                                     cell.labels = 'scGate_multi',
                                     anchor.features = hvg,
                                     dims = 1:npcs)

tree <- SampleTree.STACAS(stacas_anchors, obj.names = names(seurat.list))

integrated.ss <- IntegrateData.STACAS(anchorset=stacas_anchors, dims=npcs,
                                   sample.tree=tree)

integrated.ss <- ScaleData(integrated.ss) |> RunPCA(dims = 1:npcs) |>
      RunUMAP(reduction = "pca", dims = 1:npcs, seed.use=seed)
```

See results
```{r fig.width=10}
a <- DimPlot(integrated.ss, reduction = "umap",
        cols = mycols, group.by = "functional.cluster", raster = T) +
  ggtitle('Annotation') + theme(aspect.ratio = 1)

b <- DimPlot(integrated.ss, reduction = "umap",
        group.by = "SampleLabel", raster = T) +
  theme(aspect.ratio = 1, legend.text = element_text(size=7),
        legend.key.size = unit(0.1, 'cm'))

c <- DimPlot(integrated.ss, reduction = "umap",
        group.by = "scGate_multi", cols = mycols_scGate, raster = T) +
  theme(aspect.ratio = 1, legend.text = element_text(size=7),
        legend.key.size = unit(0.1, 'cm'))

b | a | c
```
