---
title: "CD8 TIL integration with STACAS"
author: "Paul Gueguen, Massimo Andreatta, and Santiago  Carmona"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

```{r, include=FALSE, fig.width=16, fig.height=12}
library(renv)
renv::restore()
library(Seurat)
library(ggplot2)
library(scGate)
library(STACAS)
library(SignatuR)
library(dplyr)
```

We start from integrated collection of large and high-quality samples, generated using 2-Integration_STACAS.Rmd. Here we integrate the remaining datasets on top of the integrated 'reference' set.

# Set up parameters

```{r}
path_output <- "./out/"
npcs <- 50
nfeatures <- 800
seed <- 123
set.seed(seed)
umap.neighbors <- 30
semisup=FALSE

# Setup color palette
mycols <- c('NaiveLike' = '#b7d2e0', 'CM' = '#da6f6f','EM'= '#72b28a','TEMRA' = '#e5bfaf', 'TPEX' = '#aca6e0' , 'TEX' ='#f5d39f', "MAIT" = '#fdbfd4')

mycols_scGate <- c('CD8_N' = '#b7d2e0', 'CD8_EM'= '#72b28a','CD8_TEMRA' = '#e5bfaf', 'CD8_TPEX' = '#aca6e0' , 'CD8_TEX' ='#f5d39f', "CD8_MAIT" = '#fdbfd4')
```


To limit available memory and simulate a smaller machine, we can set a memory limit in .Renviron (local for this project) to e.g. R_MAX_VSIZE=64Gb
NB! remember to comment out R_MAX_VSIZE after you're done!

# Load data

```{r}
seurat.all <- readRDS('cache/Seurat.list.Utility.v0.4_step7.rds')
length(seurat.all)

preintegrated <- readRDS('out/Utility.v0.4.CD8.integrated.stacas.semisup.rds')
```

Parameters
```{r}
npcs <- ncol(preintegrated@reductions$pca)
hvg <- VariableFeatures(preintegrated)
```

Prepare data
```{r}
inref <- unique(preintegrated$SampleLabel)
notinref <- setdiff(names(seurat.all), inref)
seurat.all <- seurat.all[notinref]

#small-scale test
#seurat.all <- seurat.all[101:150]


seurat.all <- lapply(seurat.all, NormalizeData)
seurat.all <- lapply(seurat.all, function(x){
  x$functional.cluster <- NA
  x
})

#Exclude small samples (<500 cells)
tab <- unlist(lapply(seurat.all, ncol))
pass <- names(tab)[tab > 500]
seurat.all <- seurat.all[pass]

n <- unlist(lapply(seurat.all, ncol))
head(sort(n))
```

Prepare reference sets
```{r}
#copy assay
preintegrated <- RenameAssays(preintegrated, integrated = "preintegrated")
DefaultAssay(preintegrated) <- "preintegrated"
```

# Unsupervised STACAS integration

Decide whether to use scGate labels to guide STACAS integration (cell.labels param)

```{r}
gc(reset=TRUE)
t1 <- Sys.time()

seurat.all <- append(seurat.all, values=list('reference'=preintegrated), after=0)
#obj.names <- names(seurat.all)

stacas_anchors <- FindAnchors.STACAS(seurat.all,
                                     reference = 1,
                                     cell.labels = NULL,
                                     anchor.features = hvg,
                                     dims = 1:npcs)

rm(seurat.all, preintegrated) #no longer needed, free up memory
gc()

integrated <- IntegrateData.STACAS(anchorset=stacas_anchors, dims=npcs)

#Time elapsed
t2 <- Sys.time()
t2-t1

```



Reduced dimensionalities
```{r}
integrated <- ScaleData(integrated) |> RunPCA(dims = 1:npcs) |>
      RunUMAP(reduction = "pca", dims = 1:npcs, seed.use=seed)
```

See results
```{r}
a <- DimPlot(integrated, reduction = "umap",
        cols = mycols, group.by = "functional.cluster", raster = T) +
  ggtitle('Annotation') + theme(aspect.ratio = 1)

b <- DimPlot(integrated, reduction = "umap",
        group.by = "SampleLabel", raster = T) +
  theme(aspect.ratio = 1, legend.text = element_text(size=7),
        legend.key.size = unit(0.1, 'cm'))  + NoLegend()

notna <- colnames(integrated)[!is.na(integrated$functional.cluster)]
sub <- subset(integrated, cells=notna)

c <- DimPlot(sub, reduction = "umap",
        cols = mycols, group.by = "functional.cluster", raster = T) +
  ggtitle('Annotation') + theme(aspect.ratio = 1)

b | c
```


```{r}
saveRDS(integrated, file="aux/large.integrated.unsup.rds")
```

Define a function to expand labels to anannotated cells, based on nearest neighbors
```{r}
annotate.by.neighbors <- function (object,
                                   ref.cells=NULL,  #by default use as reference all cells with a label
                                   reduction = "pca",
                                   ndim = NULL,
                                   k = 20,
                                   ncores = 1,
                                   bg.pseudocount = 10^9,
                                   labels.col = "functional.cluster") {
  
  require(BiocNeighbors)
  require(BiocParallel)
  
  if (is.null(ndim)) {
    ndim <- ncol(object@reductions[[reduction]])
  }
  
  if (is.null(ref.cells)) {
    all.labs <- object@meta.data[, labels.col]
    ref.cells <- colnames(object)[!is.na(all.labs)]
  }
  nr.cells <- setdiff(colnames(object), ref.cells)
  
  if (length(nr.cells)==0) {
    return(object)
  }
  
  ref.space <- Embeddings(object, reduction=reduction)[ref.cells, 1:ndim]
  query.space <- Embeddings(object, reduction=reduction)[nr.cells, 1:ndim]
  
  labels <- object@meta.data[ref.cells, labels.col]
  
  #calculate expected BG frequencies
  bg <- table(labels) + bg.pseudocount
  bg <- bg/(sum(bg))
  
  nn.ranked <- queryKNN(ref.space, query.space, k=k,
                        BNPARAM=AnnoyParam(),
                        BPPARAM=MulticoreParam(ncores))
  
  pred.type <- apply(nn.ranked$index, 1, function(x) {
    tab <- table(labels[x])
    tab <- tab/sum(tab)
    tab.norm <- tab
    for (i in seq_along(tab)) {
      ct <- names(tab)[i]
      tab.norm[i] <- tab[i]/bg[ct]
    }
    
    scores <- sort(tab.norm, decreasing = T)
    names(scores)[1]
  })
  
  object@meta.data[nr.cells,labels.col] <- pred.type
  return(object)
}
```

```{r}
table(integrated$functional.cluster)

integrated <- annotate.by.neighbors(integrated,
                                    k=20,
                                    labels.col = "functional.cluster",
                                    ncores=8)

table(integrated$functional.cluster)
```

```{r fig.height=8}
integrated$functional.cluster <- factor(integrated$functional.cluster,
                                        levels=names(mycols))

DimPlot(integrated, reduction = "umap",
        group.by = "functional.cluster", cols=mycols, label = T) +
  ggtitle('ssSTACAS - relabeled') + theme(aspect.ratio = 1)

ggsave("plots/umap_largescale_integration_bylabel.pdf", height=4, width=6)

```

```{r fig.height=5, fig.width=14}
DefaultAssay(integrated) <- "RNA"

integrated$functional.cluster <- factor(integrated$functional.cluster,
                                        levels=rev(names(mycols)))

Idents(integrated) <- integrated$functional.cluster
genes <- c('SELL','LEF1',"S1PR1","IL7R", "TCF7", "CCR7",
           "LMNA","GZMA","GZMK","CCL4","FGFBP2",'FCGR3A','KLRG1',"GNLY",
           'XCL1',"XCL2","GNG4","CRTAM","TOX","CXCL13",'PDCD1',
           "CXCR6","GZMB","PRF1","CTLA4","LAG3","HAVCR2",'KLRB1','SLC4A10','TRAV1-2')

v <- VlnPlot(integrated, features = genes, stack = TRUE, flip = FALSE, fill.by = "ident", cols = mycols)

ggsave("plots/large_allsamples_violins_horizontal.pdf", plot=v, height=4, width=12)

v
DefaultAssay(integrated) <- "integrated"
```

```{r fig.width=12, fig.height=4}
DefaultAssay(integrated) <- 'RNA'
p <- DotPlot(integrated, features = genes, cols="RdBu", scale=T, col.max=1.5) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

p
ggsave("plots/large_allsamples_dotplot.pdf", height=4, width=11)

#Save source data
write.csv(x=p$data, file="aux/CD8T_largescale_dotplot_source.csv")
```

## Look for conserved signal between all these datasets
```{r}
#Subset for testing
#bu <- integrated
#integrated <- subset(integrated, subset=SampleLabel %in% pass)

samples <- unique(integrated$SampleLabel)
#integrated <- subset(integrated, subset=Tissue == "Melanoma") 

RNA_log_cutoff <- 1
integrated@assays$RNA_trimmed <- integrated@assays$RNA
#integrated@assays$RNA_trimmed@data[integrated@assays$RNA_trimmed@data > RNA_log_cutoff] <- RNA_log_cutoff
```

```{r}
Idents(integrated) <- "functional.cluster"
DefaultAssay(integrated) <- "RNA_trimmed"

sort(table(integrated$functional.cluster))
mean(table(integrated$functional.cluster,integrated$SampleLabel)>100)

set.seed(1234)

marker.list <- ProjecTILs::FindAllMarkers.bygroup(integrated, split.by = "SampleLabel",
                                 min.cells.group=30, 
                                 max.cells.per.ident = 1000, 
                                 features = NULL, test.use = "wilcox",
                                 min.diff.pct=0.05, min.freq = 2/3, verbose = F)

genes <- lapply(marker.list, function(x) {
  names(x) |> head(6)
})

genes.use.bygroup.trimm <- unique(unlist(genes))
genes.use.bygroup.trimm
```

# Unsupervised heatmap

```{r fig.height=5, fig.width=7}
library(ProjecTILs)
palettes <- list()

integrated@misc$atlas.palette <- mycols

#Color palettes
palettes$Tissue <- colorRampPalette(brewer.pal(n = 8, 
            name = "Set1"))(length(unique(integrated$Tissue)))
names(palettes$Tissue) <- unique(integrated$Tissue)
palettes$SampleLabel <- colorRampPalette(brewer.pal(n = 8, 
            name = "Set2"))(length(unique(integrated$SampleLabel)))
names(palettes$SampleLabel) <- unique(integrated$SampleLabel)

##heatmap
cluster.method <- "average"
#cluster.method <- "ward.D2"
  
h <- celltype.heatmap(integrated, cluster.col = 'functional.cluster',  metadata=c("SampleLabel", "Tissue"),
                  assay = "RNA_trimmed", genes = genes.use.bygroup.trimm,
                  min.cells=30, ref=integrated, cluster_genes = T, palette = palettes,
                  cluster_samples = TRUE, method=cluster.method, scale = "none",
                  brewer.palette = "Blues", breaks = seq(0, RNA_log_cutoff, by = 0.1))

h
```