---
title: "CD8 TIL integration with STACAS"
author: "Paul Gueguen, Massimo Andreatta, and Santiago  Carmona"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

```{r, include=FALSE, fig.width=16, fig.height=12}
library(renv)
renv::restore()
library(Seurat)
library(ggplot2)
library(scGate)
library(STACAS)
library(SignatuR)
library(dplyr)
```

We start from integrated collection of large and high-quality samples, generated using 2-Integration_STACAS.Rmd. Here we integrate the remaining datasets on top of the integrated 'reference' set.

# Set up parameters

```{r}
path_output <- "./out/"
npcs <- 50
nfeatures <- 800
seed <- 123
set.seed(seed)
umap.neighbors <- 30
semisup=TRUE

# Setup color palette
mycols <- c('NaiveLike' = '#b7d2e0', 'CM' = '#da6f6f','EM'= '#72b28a','TEMRA' = '#e5bfaf', 'TPEX' = '#aca6e0' , 'TEX' ='#f5d39f', "MAIT" = '#fdbfd4')

mycols_scGate <- c('CD8_N' = '#b7d2e0', 'CD8_EM'= '#72b28a','CD8_TEMRA' = '#e5bfaf', 'CD8_TPEX' = '#aca6e0' , 'CD8_TEX' ='#f5d39f', "CD8_MAIT" = '#fdbfd4')
```

# Load data

```{r}
seurat.all <- readRDS('cache/Seurat.list.Utility.v0.4_step6.rds')
length(seurat.all)

preintegrated <- readRDS('out/Utility.v0.4.CD8.integrated.stacas.semisup.rds')
```

Parameters
```{r}
npcs <- ncol(preintegrated@reductions$pca)
hvg <- VariableFeatures(preintegrated)
```

Prepare data
```{r}
inref <- unique(preintegrated$SampleLabel)
notinref <- setdiff(names(seurat.all), inref)
seurat.all <- seurat.all[notinref]

#small-scale test
#seurat.all <- seurat.all[101:150]


seurat.all <- lapply(seurat.all, NormalizeData)
seurat.all <- lapply(seurat.all, function(x){
  x$functional.cluster <- NA
  x
})

#may want to exclude small samples
n <- unlist(lapply(seurat.all, ncol))
head(sort(n))
```

Prepare reference sets
```{r}
#copy assay
preintegrated <- RenameAssays(preintegrated, integrated = "preintegrated")
DefaultAssay(preintegrated) <- "preintegrated"
```

#Semi-supervised approach

Predict CD8T subtypes using scGate
```{r}
# Setup function to run scGate + seurat pipeline
run_seurat <- function(x, bl=NULL, nfeat=1000, npca=30) {
  
  x <- x %>% NormalizeData(verbose=F) %>%  
      FindVariableFeatures.STACAS(nfeat = nfeat, genesBlockList = bl) %>%
      ScaleData(verbose=F) %>%
      RunPCA(npcs=npca, verbose=F) %>% RunUMAP(dims=1:npca, verbose=F)

  return(x)
}

# define a function using variable object_in
run_scGate_CD8T <- function(x, models, reduction="pca") {

  x <- scGate(x, model=models, multi.asNA=T, ncores=8,
           reduction=reduction, assay = "RNA")

  return(x)
}
```

```{r message=F, results=F}
#scGate_models <- scGate::get_scGateDB(branch = "dev")
scGate_models <- scGate::get_scGateDB(version = 'v0.12') #fix version for reproducibility

cd8.til.models <- scGate_models$human$CD8_TIL
cd8.til.models <- cd8.til.models[names(cd8.til.models) %in% c("CD8_Tinn","CD8_TRM") == FALSE]

my.genes.blocklist <- SignatuR::GetSignature(SignatuR$Hs)

names <- names(seurat.all)
seurat.all <- lapply(
  seq_along(seurat.all), 
  function(i) {
    print(sprintf("Dataset %i of %i", i, length(seurat.all)))
    x <- seurat.all[[i]]
    x <- run_seurat(x, bl=my.genes.blocklist, nfeat=1000, npca=30)
    run_scGate_CD8T(x, models=cd8.til.models)
  })
names(seurat.all) <- names

#Also run on reference objects
preintegrated <- run_scGate_CD8T(preintegrated, models=cd8.til.models)
```

```{r}
DefaultAssay(preintegrated) <- "preintegrated"
ref <- list('reference'=preintegrated)
seurat.list <- append(seurat.all, values=ref, after=0)


head(sapply(seurat.list, DefaultAssay))

stacas_anchors <- FindAnchors.STACAS(seurat.list,
                                     reference = 1,
                                     cell.labels = 'scGate_multi',
                                     anchor.features = hvg,
                                     dims = 1:npcs)

tree <- SampleTree.STACAS(stacas_anchors, obj.names = names(seurat.list))

integrated.ss <- IntegrateData.STACAS(anchorset=stacas_anchors, dims=npcs,
                                   sample.tree=tree)

integrated.ss <- ScaleData(integrated.ss) |> RunPCA(dims = 1:npcs) |>
      RunUMAP(reduction = "pca", dims = 1:npcs, seed.use=seed)
```

Free up some space
```{r}
rm(seurat.all)
gc()
```

```{r}
saveRDS(integrated.ss, file="aux/large.integrated.ss.rds")
```

See results
```{r fig.width=10}
a <- DimPlot(integrated.ss, reduction = "umap",
        cols = mycols, group.by = "functional.cluster", raster = T) +
  ggtitle('Annotation') + theme(aspect.ratio = 1)

b <- DimPlot(integrated.ss, reduction = "umap",
        group.by = "SampleLabel", raster = T) +
  theme(aspect.ratio = 1, legend.text = element_text(size=7),
        legend.key.size = unit(0.1, 'cm')) + NoLegend()

notna <- colnames(integrated.ss)[!is.na(integrated.ss$functional.cluster)]
sub <- subset(integrated.ss, cells=notna)

c <- DimPlot(sub, reduction = "umap",
        cols = mycols, group.by = "functional.cluster", raster = T) +
  ggtitle('Annotation') + theme(aspect.ratio = 1)

b | c
```

Define a function to expand labels to anannotated cells, based on nearest neighbors
```{r}
annotate.by.neighbors <- function (object,
                                   ref.cells=NULL,  #by default use as reference all cells with a label
                                   reduction = "pca",
                                   ndim = NULL,
                                   k = 20,
                                   ncores = 1,
                                   labels.col = "functional.cluster") {
  
  require(BiocNeighbors)
  require(BiocParallel)
  
  if (is.null(ndim)) {
    ndim <- ncol(object@reductions[[reduction]])
  }
  
  if (is.null(ref.cells)) {
    all.labs <- object@meta.data[, labels.col]
    ref.cells <- colnames(object)[!is.na(all.labs)]
  }
  nr.cells <- setdiff(colnames(object), ref.cells)
  
  if (length(nr.cells)==0) {
    return(object)
  }
  
  ref.space <- Embeddings(object, reduction=reduction)[ref.cells, 1:ndim]
  query.space <- Embeddings(object, reduction=reduction)[nr.cells, 1:ndim]
  
  labels <- object@meta.data[ref.cells, labels.col]
  
  nn.ranked <- queryKNN(ref.space, query.space, k=k,
                        BNPARAM=AnnoyParam(),
                        BPPARAM=MulticoreParam(ncores))
  
  pred.type <- apply(nn.ranked$index, 1, function(x) {
    scores <- sort(table(labels[x], useNA = "ifany"), decreasing = T)/k
    names(scores)[1]
  })
  
  object@meta.data[nr.cells,labels.col] <- pred.type
  return(object)
}
```

```{r}
integrated.ss <- annotate.by.neighbors(integrated.ss,
        #                            ref.cells = ref.cells,
                                    k=20,
                                    labels.col = "functional.cluster",
                                    ncores=8)

table(integrated.ss$functional.cluster)
```


```{r fig.height=8, fig.width=5}
DefaultAssay(integrated.ss) <- "RNA"
Idents(integrated.ss) <- integrated.ss$functional.cluster
genes <- c('SELL','LEF1', "TCF7", "CCR7","LMNA","IL7R","GZMK","CCL4",
           "FGFBP2",'FCGR3A','XCL1',"XCL2",'GNG4',
           "CRTAM","TOX",'PDCD1',"HAVCR2","LAG3","GZMB","PRF1",
           'KLRB1','TRAV1-2','SLC4A10')
VlnPlot(integrated.ss, pt.size = 0.1, features = genes, cols=mycols,
        stack = TRUE, flip = TRUE, assay = "RNA", fill.by = "ident")
DefaultAssay(integrated.ss) <- "integrated"
```
